---
title: "MKT615: Data Storytelling for Marketers"
subtitle: "Lecture 5.3: Data Visualization"
author: "Davide Proserpio"
institute: "Marshall School of Business"
# date: "08/25/2021" #(updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      beforeInit: "libs/cols_macro.js"
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
      ratio: "16:9"

---

```{css, echo=FALSE}
@media print {
  .has-continuation {
    display: block !important;
  }
}
.large4 { font-size: 400% }
.large2 { font-size: 200% }
.small90 { font-size: 90% }
.small75 { font-size: 75% }
```

```{r setup, include=FALSE}
library(hrbrthemes)
library(fontawesome)
library(xaringan)
library(ggplot2)
library(gganimate)
library(ggthemes)
library(gapminder)
library(tidyverse)
options(htmltools.dir.version = FALSE)

library(knitr)
opts_chunk$set(
  fig.align="center", #fig.width=6, fig.height=4.5, 
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=F#, echo=F, warning=F, message=F
  )
```
# Table of contents

1. [Checklist](#check)
2. [ggplot2](#ggplot2)
3. [Aesthetic](#aesthetic)
4. [Geoms](#geoms)
5. [Layers](#layers)
6. [More things](#more)
7. [Exercises](#exe)
8. [Readings](#readings)

---
name: check

# Checklist

We will need **ggplot2**, **gapminder**, **dplyr**, **data.table**, **ggthemes**, **gganimate**, **hrbrthemes**, **scales**, **tidyverse**

---
class: inverse, center, middle
name:ggplot2

# ggplot2

<!-- <html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html> -->


---

# Elements of ggplot2

[Hadley Wickham's](http://hadley.nz/) ggplot2 is one of the most popular R packages. 
- It also happens to be built upon some deep visualization theory: i.e. Leland Wilkinson's [*The Grammar of Graphics*](https://www.amazon.com/Grammar-Graphics-Statistics-Computing/dp/0387245448).

There's a lot to say about ggplot2's implementation of this "grammar of graphics" approach, but the three key elements are:

1. Your plot ("the visualization") is linked to your variables ("the data") through various **aesthetic mappings**.

2. Once the aesthetic mappings are defined, you can represent your data in different ways by choosing different **geoms** (i.e. "geometric objects" like points, lines or bars).

3. You build your plot in **layers**.

--

Let's review each element in turn with some actual plots.

---
class: inverse, center, middle
name: aesthetic

# Aesthetics

<!-- <html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html> -->

---

# 1. Aesthetic mappings

```{r, dev = 'svg', fig.width=6,fig.height=4}
p = ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) 
p = p + geom_point()
p
```


---

# 1. Aesthetic mappings (cont.)

```{r aesthetics1, eval=FALSE}
p = ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) 
p = p + geom_point() 
p
```

Focus on the top line, which contains the initialising `ggplot()` function call. This function accepts various arguments, including:
- Where the data come from (i.e. `data = gapminder`).
- What the aesthetic mappings are (i.e. `mapping = aes(x = gdpPercap, y = lifeExp)`).
- The `aes()` function gathers together each of the aesthetic mappings used by a layer and passes them to the layer’s mapping argument.

--

In this case, the aesthetic mappings here are pretty simple: They just define an x-axis (GDP per capita) and a y-axis (life expecancy).
- However, aesthetics are a very powerful tool. To get a sense of the power and flexibility that comes with this approach, consider what happens if we add more aesthetics to the plot call.

---

# 1. Aesthetic mappings (cont.)

E.g., let's map **color** and point **size** to two variables:

```{r aesthetics2, dev='svg', fig.width=6, fig.height=3.3}
p = ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, size = pop, color = continent))
p = p + geom_point()
p
```

--

Note that I've dropped the "data =" and "mapping =" part of the ggplot call (`ggplot2` knows the order of the arguments.)

---

# 1. Aesthetic mappings (cont.)

E.g., let's map **color** and point **size** to two variables:

```{r aesthetics22, dev='svg', fig.width=6, fig.height=3.7, echo = F}
p = ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, size = pop, color = continent))
p = p + geom_point()
p
```

- As you can see, ggplot automatically assigns *levels* to the aesthetics and creates a legend for each of them. 
- (For `x` and `y` aesthetics, ggplot2 does not create a legend, but it creates an axis line with tick marks and a label.)

---

# 1. Aesthetic mappings (cont.)

We can specify aesthetic mappings in the geom layer too. Doing so aesthetics are applied only locally (i.e., to the specific geometry).
```{r aesthetics3, dev='svg', fig.width=6, fig.height=3.3}
p = ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) ## Applicable to all geoms
p = p + geom_point(aes(size = pop, col = continent)) ## Applicable to this geom only
p
```

---

# 1. Aesthetic mappings (cont.)

Oops. What went wrong here?
```{r aesthetics4, dev='svg', fig.width=5, fig.height=3}
p = ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
p = p + geom_point(aes(size = "big", col="black"))
p
```

--

**Answer: **Aesthetics must be mapped to variables, not descriptions!

---


# 1. Aesthetic mappings (cont.)

Note that outside of the `aes()` function, you can pass values and set the aesthetic **manually**, e.g.,

```{r aesthetics44, dev='svg', fig.width=5, fig.height=3}
p = ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
p = p + geom_point(size = 0.1, col = "blue")
p
```

---
class: inverse, center, middle
name: geoms

# Geoms

<!-- <html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html> -->

---

# 2. Geoms

Once your variable relationships have been defined by the aesthetic mappings, you can invoke and combine different geoms to generate different visualizations.

```{r geoms1, dev='svg', fig.width=5, fig.height=3}
p = ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
p = p + geom_point()
p = p + geom_smooth(method = "loess")
p
```

---

# 2. Geoms (cont.)

Aesthetics can be applied differentially across geoms.

```{r geoms2, dev='svg', fig.width=5, fig.height=3}
p = ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
p = p + geom_point(aes(size = pop, col = continent))
p = p + geom_smooth(method = "loess")
p
```

---

# 2. Geoms (cont.)

The previous plot provides a good illustration of the power (or effect) that comes from assigning aesthetic mappings "globally" vs "locally" (i.e., in the individual geom layers.)
- Compare: What happens if you run the below code chunk?

```{r geoms3, eval=F, fig.width=5, fig.height=3}
p = ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, size = pop, col = continent))
p = p + geom_point(alpha = 0.3)
p = p + geom_smooth(method = "loess")
p
```

---

# 2. Geoms (cont.)

You can use the same idea to specify different data for each layer.

```{r geoms6, dev='svg', fig.width=6, fig.height=3.5}
p = ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
p = p + geom_point(aes(color = year))
p = p + geom_smooth( data = subset(gapminder, year == 2002), se = F)
p
```


---

# 2. Geoms (cont.)

Similarly, note that some geoms only accept a subset of mappings. E.g. `geom_density()` doesn't know what to do with the "y" aesthetic mapping.

```{r geoms4, error=TRUE, dev='svg'}
p = ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
p + geom_density()
```

---

# 2. Geoms (cont.)

We can fix that by being more careful about how we build the plot.

```{r geoms5, dev='svg', fig.width=6, fig.height=3.5}
p = ggplot(data = gapminder) ## i.e. No "global" aesthetic mappings"
p + geom_density(aes(x = gdpPercap, fill = continent), alpha=0.3)
```



---
# 2. Geoms (cont.)

- ggplot2 provides over 40 geoms, and extension packages provide even more (see https://exts.ggplot2.tidyverse.org/gallery/ for a sampling).
- The best way to get a comprehensive overview is the ggplot2 cheatsheet, which you can find at http://rstudio.com/resources/cheatsheets. 
- To learn more about any single geom, use help: ?geom_smooth.

---
class: inverse, center, middle
name: build

# Layers

<!-- <html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html> -->

---

# 3. Build your plot in layers

We've already seen how we can chain (or "layer") consecutive plot elements.

But it bears repeating: You can build out some truly impressive complexity and transformation of your visualization through this simple layering process.
- You don't have to transform your original data; ggplot2 takes care of all of that.
- For example (see next slide for figure).

```{r layers1}
p2 =
  ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(size = pop, col = continent), alpha = 0.3) +
  scale_color_brewer(name = "Continent", palette = "Set1") + ## Different color scale
  scale_size(name = "Population", labels = scales::comma) + ## Different point (i.e. legend) scale
  scale_x_log10(labels = scales::dollar) + ## Switch to logarithmic scale on x-axis. Use dollar units.
  labs(x = "\n Log (GDP per capita)", y = "Life Expectancy\n") + ## Better axis titles
  theme_minimal() ## Try a minimal (b&w) plot theme
```

---

# 3. Build your plot in layers (cont.)

```{r layers2, dependson=p2, echo=FALSE, dev='svg', fig.width=6.5, fig.height=4}
p2
```

---
class: inverse, center, middle
name: more

# More things

<!-- <html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html> -->


---

# Facets

One way to add additional variables is with aesthetics. Another way, particularly useful for categorical variables, is to split your plot into **facets**, subplots that each display one subset of the data.

```{r facet, dev='svg', fig.width=5, fig.height=3}
p = ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
p = p + geom_point()
p = p + facet_wrap(~ continent, nrow = 2)
p
```

---
# Facets (cont.)

To facet your plot on the combination of two variables, add facet_grid() to your plot call.

```{r facetgrid, dev='svg', fig.width=5, fig.height=3}
p = ggplot(data = subset(gapminder, year %in% c(2002,2007)), aes(x = gdpPercap, y = lifeExp))
p = p + geom_point()
p = p + facet_grid( year ~ continent)
p
```

---

# Facets (cont.)

If you prefer to not facet in the rows or columns dimension, use a . instead of a variable name, e.g. `+ facet_grid(. ~ cyl)` (the dot is optional).

```{r facetgrid2, dev='svg', fig.width=5, fig.height=3}
p = ggplot(data = subset(gapminder, year %in% c(2002,2007)), aes(x = gdpPercap, y = lifeExp))
p = p + geom_point()
p = p + facet_grid( ~ continent)
p
```

---

# Statistical transformations

`stat_summary` is very useful to create plots with summary stats w/o altering the original data

```{r summary, dev='svg', fig.width=5, fig.height=3}
p = ggplot(data = gapminder, aes(x = year))
p = p + stat_summary(aes(y=gdpPercap), fun = "mean", geom = "point")
p
```
--

In the exercises, we are going to see a few additional ways to do statistical transformations.

---


# Saving plots

`ggsave` is very useful to save plots in any format. Note that for papers, the best format is pdf.

```{r save, dev='svg', fig.width=5, fig.height=3, fig.show='hide'}
p = ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
p = p + geom_point()
p
ggsave("figures/test.pdf", p, h = 3, w = 5)
```

--

Note that you can also combine more than one plot in one pdf, but you need the `patchwork` library:

```{r save2, dev='svg', fig.width=5, fig.height=3, fig.show='hide'}
library(patchwork)
p = ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
p = p + geom_point()
p1 = p
p2 = p
ggsave("figures/side-by-side.pdf", p1+p2, h = 3, w = 5)
```



---

# There is much more

We have barely scratched the surface of ggplot2's functionality... let alone talked about the entire ecosystem of packages that has been built around it.
- Here's are two quick additional examples

--

Note that you will need to install and load some additional packages if you want to recreate the next two figures on your own machine. A quick way to do this:

```{r pacinstall}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(hrbrthemes, gganimate)
```

---

# There is much more (cont.)

Simple extension: Use an external package theme.
```{r modern_rc_theme, dependson=p2, dev='svg', fig.width=6.5, fig.height=4}
# library(hrbrthemes)
p2 + theme_modern_rc() + geom_point(aes(size = pop, col = continent), alpha = 0.2)
```

---

# There is much more (cont.)

Elaborate extension: Animation!
```{r gganim1, eval=FALSE}
# library(gganimate)
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  ease_aes('linear')
```

---

<!-- --- -->

<!-- #More (cont.) -->

<!-- ```{r ggamin2, echo=FALSE, eval=TRUE} -->
<!-- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) + -->
<!--   geom_point(alpha = 0.7, show.legend = FALSE) + -->
<!--   scale_colour_manual(values = country_colors) + -->
<!--   scale_size(range = c(2, 12)) + -->
<!--   scale_x_log10(labels = scales::dollar) + -->
<!--   facet_wrap(~continent) + -->
<!--   # Here comes the gganimate specific bits -->
<!--   labs(title = 'Year: {frame_time}', x = 'Log (GDP per capita)', y = 'Life expectancy') + -->
<!--   transition_time(year) + -->
<!--   ease_aes('linear') -->
<!-- ``` -->

name: exe

# Exercises 

Let's use the ggplot2 `mpg` dataset. What does it contain?

1. Plot the average highway miles per gallon (a dot) by type of car (class)
  <!-- p = ggplot(mpg, aes(x=class)) -->
  <!-- p = p + stat_summary(aes(y = hwy), fun = 'mean', geom = 'point') -->
  <!-- p -->

2. Rename x-axis label to "Type of Car", and y-axis label to "Hwy Miles/Gallon"

3. Create a bar plot of the number of cars by type of car (class)
  <!-- p = ggplot(mpg, aes(x = class)) -->
  <!-- p = p + geom_bar() -->
  <!-- p -->
4. Reorder the plot above by count (highest bar first)
  <!-- p = ggplot(mpg, aes(x = fct_infreq(class))) -->
  <!-- p = p + geom_bar() -->
  <!-- p -->

5. Create a scatter plot of highway per miles against (x_axis) engine displacement, color the dots by type of car (class)
  <!-- ggplot(data = mpg) +  -->
  <!-- geom_point(mapping = aes(x = displ, y = hwy, color = class)) -->

6. Do the plot above by create a panel for each type of car (use facets)
  <!-- ggplot(data = mpg) +  -->
  <!-- geom_point(mapping = aes(x = displ, y = hwy)) +  -->
  <!-- facet_wrap(~ class, nrow = 2) -->

---
name: readings

# Readings

I want you to do some reading and practice on your own. Pick either of the following (or choose among the litany of online resources) and work through their examples:
- [Chapter 3](https://r4ds.had.co.nz/data-visualisation.html) of *R for Data Science* by Hadley Wickham and Garett Grolemund.
- [*Data Visualization: A Practical Guide*](https://socviz.co/makeplot.html) by Kieran Healy.
- [Designing ggplots](https://designing-ggplots.netlify.com) by Malcom Barrett.

