---
title: "MKT615: Data Storytelling for Marketers"
subtitle: "Lecture 5.3: Data Visualization"
author: "Davide Proserpio"
institute: "Marshall School of Business"
# date: "08/25/2021" #(updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      beforeInit: "libs/cols_macro.js"
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
      ratio: "16:9"

---

```{css, echo=FALSE}
@media print {
  .has-continuation {
    display: block !important;
  }
}
.large4 { font-size: 400% }
.large2 { font-size: 200% }
.small90 { font-size: 90% }
.small75 { font-size: 75% }
```

```{r setup, include=FALSE}
library(hrbrthemes)
library(fontawesome)
library(xaringan)
options(htmltools.dir.version = FALSE)

library(knitr)
opts_chunk$set(
  fig.align="center", #fig.width=6, fig.height=4.5, 
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=F#, echo=F, warning=F, message=F
  )
```
# Table of contents

1. [Checklist](#check)
2. [Introduction](#intro)


---

---

# Install and load (cont.)

Once the packages are installed, load them into your R session with the `library()` function.
```{r pkg_load}
library(ggplot2)
library(gapminder) 
```

---

# Elements of ggplot2

[Hadley Wickham's](http://hadley.nz/) ggplot2 is one of the most popular packages in the entire R canon. 
- It also happens to be built upon some deep visualization theory: i.e. Leland Wilkinson's [*The Grammar of Graphics*](https://www.amazon.com/Grammar-Graphics-Statistics-Computing/dp/0387245448).

There's a lot to say about ggplot2's implementation of this "grammar of graphics" approach, but the three key elements are:

1. Your plot ("the visualization") is linked to your variables ("the data") through various **aesthetic mappings**.

2. Once the aesthetic mappings are defined, you can represent your data in different ways by choosing different **geoms** (i.e. "geometric objects" like points, lines or bars).

3. You build your plot in **layers**.

--

</br>

That's kind of abstract. Let's review each element in turn with some actual plots.

---

# 1. Aesthetic mappings

```{r, dev = 'svg', fig.width=6,fig.height=4}
p = ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) 
p = p + geom_point(size = 0.2)
p
```


---

# 1. Aesthetic mappings (cont.)

```{r aesthetics1, eval=FALSE}
p = ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) 
p = p + geom_point(size = 0.2) #size of the geom points
p
```

Focus on the top line, which contains the initialising `ggplot()` function call. This function accepts various arguments, including:
- Where the data come from (i.e. `data = gapminder`).
- What the aesthetic mappings are (i.e. `mapping = aes(x = gdpPercap, y = lifeExp)`).

--

The aesthetic mappings here are pretty simple: They just define an x-axis (GDP per capita) and a y-axis (life expecancy).
- To get a sense of the power and flexibility that comes with this approach, however, consider what happens if we add more aesthetics to the plot call...

---

# 1. Aesthetic mappings (cont.)

```{r aesthetics2, dev='svg', fig.width=6, fig.height=3.7}
p = ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, size = pop, col = continent))
p = p + geom_point(alpha = 0.3) ## "alpha" controls transparency. Takes a value between 0 and 1.
p
```

--

Note that I've dropped the "data =" and "mapping =" part of the ggplot call (`ggplot2` knows the order of the arguments.)

---

# 1. Aesthetic mappings (cont.)

We can specify aesthetic mappings in the geom layer too.
```{r aesthetics3, dev='svg', fig.width=6, fig.height=3.7}
p = ggplot(gapminder, aes(x = gdpPercap, y = lifeExp)) ## Applicable to all geoms
p = p + geom_point(aes(size = pop, col = continent), alpha = 0.3) ## Applicable to this geom only
p
```

---

# 1. Aesthetic mappings (cont.)

Oops. What went wrong here?
```{r aesthetics4, dev='svg', fig.width=5, fig.height=3}
p = ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
p = p + geom_point(aes(size = "big", col="black"), alpha = 0.3)
p
```

--

**Answer: **Aesthetics must be mapped to variables, not descriptions!

---

# 2. Geoms

Once your variable relationships have been defined by the aesthetic mappings, you can invoke and combine different geoms to generate different visulaizations.

```{r geoms1, dev='svg', fig.width=5, fig.height=3}
p = ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
p = p + geom_point(alpha = 0.3)
p = p + geom_smooth(method = "loess")
p
```

---

# 2. Geoms (cont.)

Aesthetics can be applied differentially across geoms.

```{r geoms2, dev='svg', fig.width=5, fig.height=3}
p = ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
p = p + geom_point(aes(size = pop, col = continent), alpha = 0.3)
p = p + geom_smooth(method = "loess")
p
```

---

# 2. Geoms (cont.)

The previous plot provides a good illustration of the power (or effect) that comes from assigning aesthetic mappings "globally" vs in the individual geom layers.
- Compare: What happens if you run the below code chunk?

```{r geoms3, eval=F, fig.width=5, fig.height=3}
p = ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, size = pop, col = continent))
p = p +geom_point(alpha = 0.3)
p = p + geom_smooth(method = "loess")
p
```

---

# 2. Geoms (cont.)

Similarly, note that some geoms only accept a subset of mappings. E.g. `geom_density()` doesn't know what to do with the "y" aesthetic mapping.

```{r geoms4, error=TRUE, dev='svg'}
p = ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp))
p + geom_density()
```

---

# 2. Geoms (cont.)

We can fix that by being more careful about how we build the plot.

```{r geoms5, dev='svg', fig.width=6, fig.height=3.5}
p = ggplot(data = gapminder) ## i.e. No "global" aesthetic mappings"
p + geom_density(aes(x = gdpPercap, fill = continent), alpha=0.3)
```

---

# 3. Build your plot in layers

We've already seen how we can chain (or "layer") consecutive plot elements.

But it bears repeating: You can build out some truly impressive complexity and transformation of your visualization through this simple layering process.
- You don't have to transform your original data; ggplot2 takes care of all of that.
- For example (see next slide for figure).

```{r layers1}
p2 =
  ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp)) +
  geom_point(aes(size = pop, col = continent), alpha = 0.3) +
  scale_color_brewer(name = "Continent", palette = "Set1") + ## Different color scale
  scale_size(name = "Population", labels = scales::comma) + ## Different point (i.e. legend) scale
  scale_x_log10(labels = scales::dollar) + ## Switch to logarithmic scale on x-axis. Use dollar units.
  labs(x = "\n Log (GDP per capita)", y = "Life Expectancy\n") + ## Better axis titles
  theme_minimal() ## Try a minimal (b&w) plot theme
```

---

# 3. Build your plot in layers (cont.)

```{r layers2, dependson=p2, echo=FALSE, dev='svg', fig.width=6.5, fig.height=4}
p2
```

---

# What else?

We have barely scratched the surface of ggplot2's functionality... let alone talked about the entire ecosystem of packages that has been built around it.
- Here's are two quick additional examples

--

Note that you will need to install and load some additional packages if you want to recreate the next two figures on your own machine. A quick way to do this:

```{r pacinstall}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(hrbrthemes, gganimate)
```

---

# What else? (cont.)

Simple extension: Use an external package theme.
```{r modern_rc_theme, dependson=p2, dev='svg', fig.width=6.5, fig.height=4}
# library(hrbrthemes)
p2 + theme_modern_rc() + geom_point(aes(size = pop, col = continent), alpha = 0.2)
```

---

# What else? (cont.)

Elaborate extension: Animation! (See the next slide for the resulting GIF.)
```{r gganim1, eval=FALSE}
# library(gganimate)
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  ease_aes('linear')
```

---

# What else? (cont.)

```{r ggamin2, echo=FALSE}
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10(labels = scales::dollar) +
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'Log (GDP per capita)', y = 'Life expectancy') +
  transition_time(year) +
  ease_aes('linear')
```

--

<!-- Note that this animated plot provides a much more intuitive understanding of the underlying data. Just as [Hans Rosling](https://www.ted.com/talks/hans_rosling_the_best_stats_you_ve_ever_seen) intended. -->

---
# Readings

I want you to do some reading and practice on your own. Pick either of the following (or choose among the litany of online resources) and work through their examples:
- [Chapter 3](https://r4ds.had.co.nz/data-visualisation.html) of *R for Data Science* by Hadley Wickham and Garett Grolemund.
- [*Data Visualization: A Practical Guide*](https://socviz.co/makeplot.html) by Kieran Healy.
- [Designing ggplots](https://designing-ggplots.netlify.com) by Malcom Barrett.

