---
title: "MKT615: Data Storytelling for Marketers"
subtitle: "Lecture 1: Introduction"
author: "Davide Proserpio"
institute: "Marshall School of Business"
date: "08/25/2021" #(updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---
```{r setup, include=FALSE}
library(tidyverse)
library(hrbrthemes)
library(fontawesome)
```

# Introduction

### About me

`r fa('graduation-cap')` Assistant Professor of Marketing

`r fa('address-book')` [Davide Proserpio](https://dadepro.github.io/)

`r fa('envelope')` [proserpi@marshall.usc.edu](proserpi@marshall.usc.edu)


### About you

Names, what you are interest in, etc.

---

class: inverse, center, middle

 # Course overview

---

# Course: Data Storytelling for Marketing

I will teach the first seven weeks (then Dinesh will take over)

--

All the material is available at https://github.com/dadepro/mkt-615

We are going to use Git and GitHub to download course material and upload homeworks, assigments, and presentations

--

What will we learn?

- We will focus on practical skill and tools that will benfit your research process
  1. Project management
  2. Version control (Git and GitHub)
  3. Accessing computing resources at USC (Shell)
  3. Data manipulation, cleaning, visualization (R)
  4. Data storage (MySQL)

--

> *"In short, we will cover things that I wish someone had taught me when I was starting out in graduate
school."*

--

Please bring your laptops to class. This will be a very hands-on course. We'll be working through lecture notes together in class and youll be running code on your own laptops

---

# Grading (first half only)

| Component                           | Weight |
|:-|-:|
| 2 × homework assignments (25% each) | 50% |
| 1 × project 		| 40% |
| 1 x short presentation | 10% |

The project involves the analysis of a dataset of your choice using R and Rstudio
- You will present your data analysis on **Ocober 6**
- Where can I find datasets? 
  - [Google dataset search engine](https://datasetsearch.research.google.com/)
  - [Kaggle](https://www.kaggle.com/competitions)
  - All Marketing Science papers come with replication code and data
  - You can scrape it yourself (advanced, we won't have time to cover scraping)
  - Me (Airbnb, Yelp, Expedia)

- Short presentation of a R package related to the lecture (I will provide a list among which to choose)

---
class: inverse, center, middle

 # Questions?

---
# Setting up your laptops for the class

### Mshresearch (Server that I manage and that you can use)
Characteristics: Ubuntu OS, 64 cores, 0.5TB of RAM

All of you should have received an email to setup your mshresearch.marshall.usc account

Software installed
- Rstudio sever: http://mshresearch.marshall.usc.edu:8787/ (you can access with the mshresearch user and pwd)
- Stata 16 
- Matlab 2021
- Python 3.7
- MySQL

---

# Setting up your laptops for the class (cont.)

### R and Rstudio
Download the latest version of [R](https://www.r-project.org/) and [RStudio](https://www.rstudio.com/products/rstudio/)

Installing useful packages:
```{r eval=FALSE, cache=FALSE}
# Package names
packages <- c("sjmisc", "RMySQL", "stargazer", "fixest", "ggthemes" , "zoo", "data.table", "lfe", "ggplot2", "readxl", "dplyr", "tidyr", "ggfortify", 
              "DT", "reshape2", "knitr", "lubridate", "pwr", "psy", "car", "doBy", "imputeMissings", "RcmdrMisc", "questionr", 
              "vcd", "multcomp", "KappaGUI", "rcompanion", "FactoMineR", "factoextra", "corrplot", "ltm", "goeveg", "corrplot", 
              "FSA", "MASS", "scales", "nlme", "psych", "ordinal", "lmtest", "ggpubr", "dslabs", "stringr", "assist", "ggstatsplot", 
              "forcats", "styler", "remedy", "addinslist", "esquisse", "here", "summarytools", "magrittr", "tidyverse", 
              "funModeling", "pander", "cluster", "abind")

# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}
```

---

# Setting up your laptops for the class (cont.)

### MySQL Client

**Mac:** [Sequel Pro](https://www.sequelpro.com/)

**Windows:** [MySQL Workbench](https://www.mysql.com/products/workbench/)

---

# Setting up your laptops for the class (cont.)

### Git and GitHub

Download [Git](https://git-scm.com/downloads) (Windows users might also need [Git for Windows](https://gitforwindows.org/) so that they have a Bash shell)

Create a [GitHub](https://github.com/) account

Let me know if you need help, one place to start is https://happygitwithr.com/

---

# Setting up your laptops for the class (cont.)

### Some OS-specific extras

- **Windows:** Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/). I also recommend that you install [Chocolately](https://chocolatey.org/).

- **Mac:** Install [Homebrew](https://brew.sh/). I also recommend that you configure/open your C++ toolchain (see [here](https://github.com/rmacoslib/r-macos-rtools#installer-package-for-macos-r-toolchain-).)

- **Linux:** None (you should be good to go).

---

# Checklist

1. Do you have the most recent version of R and Rstudio?
  - Check R version from RStudio: 
      
      `version$version.string`
  
  - Check RStudio version: 
    
      `RStudio.Version()$version`
  
2. Update all R Packages: 
      
      `update.packages(ask = FALSE, checkBuilt = TRUE)`
      
---

# Checklist (cont.)

Open the Bash [Shell](https://happygitwithr.com/shell.html#shell)
- Check Git version: `git --version`
  
- Configure Git information (substitue in your details):
  ```{r eval=FALSE, cache=FALSE}
  git config --global user.name 'Davide Proserpio'
  git config --global user.email 'proserpi@marshall.usc.edu'
  git config --global --list
  ```
  
  or 
  
  `git config --global --edit` and use [vim](https://www.vim.org/) to modify the file

- Did you create a GitHub account?

- Recently GitHub updated its security requirements and now uses 2FA. This means that to access Git via Shell you need a token
  - Create a [GitHub access token](https://docs.github.com/en/github/authenticating-to-github/keeping-your-account-and-data-secure/creating-a-personal-access-token)
  - **Save it somewhere**, you'll need it to authenticate using the Shell