---
title: "MKT615: Data Storytelling for Marketers"
subtitle: "Lecture 4: R Basics"
author: "Davide Proserpio"
institute: "Marshall School of Business"
# date: "08/25/2021" #(updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
      ratio: "16:9"

---

```{r setup, include=FALSE}
library(tidyverse)
library(hrbrthemes)
library(fontawesome)
library(xaringan)
options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align="center", #fig.width=6, fig.height=4.5, 
  # out.width="748px", #out.length="520.75px",
  dpi=300, #fig.path='Figs/',
  cache=F#, echo=F, warning=F, message=F
  )
```

class: inverse, center, middle
name: intro

# Introduction to R
<!-- <html><div style='float:left'></div><hr color='#EB811B' size=1px width=796px></html> -->

---

# Agenda

This lecture is going to be very hands on to get you comfortable with R, so launch RStudio and be prepared to code

<!-- ### Packages that you will need for today. -->

We're going to work almost exclusively in **base** R today but very briefly to learn some useful notations and operators. What I want you to learn very well is **tidyverse** and **data.table** which are the subjects of the next two lectures.

<!-- - I'll also use the [**dplyr**](https://dplyr.tidyverse.org/) package, but only to demonstrate a few additional considerations for working with non-base libraries. Install/update it now, either through RStudio (recommended) or from your R console (`install.packages("dplyr"), dependencies = TRUE`).  -->

<!-- - If it is already installed, just load it:  -->
<!-- `library(dplyr)` -->

---

# Some useful operators

- Not or Negate: `!`

- Value matching: `%in%`

- Evaluation (of equality): `==`

- Assignment: either `<-` or `=` 
  - I use `=`, you can use whichever you prefer but be consistent
  
- Find missing values (`NA` in R): `is.na()`

- Help with commands/functions: `?` followed by the command/function name

- Commenting code: `#`

- Code examples: `example()`

- Vignettes: `vignette()` 

--

We are going to try to use all these operators today.
---

# Moere about R

- It is an object [object-oriented programming language](https://en.wikipedia.org/wiki/Object-oriented_programming)

- Everything is an **object**

### What are objects? 

It's important to emphasize that there are many different *types* (or *classes*) of objects. 

It is helpful simply to name some objects that we'll be working with regularly:
- vectors
- matrices
- data frames
- data tables
- lists
- functions
- etc.

---

# What are objects? (cont.)
Each object class has its own set of rules ("methods") for determining valid operations
- For example, you can perform many of the same operations on matrices and data frames, but not all
- At the same time, you can (usually) convert an object from one type to another.

```{r d}
## Create a small data frame called "d".
d = data.frame(x = 1:2, y = 3:4) 
d
```

```{r m}
## Convert it to (i.e. create) a matrix call "m".
m = as.matrix(d)
m
```

---

# Object class, type, and structure

Use the `class`, `typeof`, and `str` commands if you want understand more about a particular object.

```{r}
# d = data.frame(x = 1:2, y = 3:4) ## Create a small data frame called "d".
class(d) ## Evaluate its class.
typeof(d) ## Evaluate its type.
str(d) ## Show its structure.
```

--

PS: Confused by the fact that `typeof(d)` returns "list"? See [here](https://stackoverflow.com/questions/45396538/typeofdata-frame-shows-list-in-r).

---

# Object class, type, and structure (cont.)

Of course, you can always just inspect/print an object directly in the console.
- E.g. Type `d` and hit Enter.


The `View()` function is also very helpful. This is the same as clicking on the object in your RStudio *Environment* pane. (Try both methods now.)
- E.g. `View(d)`.

---

# Reserved words

We've seen that we can assign objects to different names. However, there are a number of special words that are "reserved" in R.
- These are are fundamental commands, operators and relations in base R that you cannot (re)assign, even if you wanted to.
- We already encountered examples with the logical operators.

See [here](http://stat.ethz.ch/R-manual/R-devel/library/base/html/Reserved.html) for a full list, including (but not limited to):

```R
if 
else 
while 
function 
for
TRUE 
FALSE 
NULL 
Inf 
NaN 
NA 
```

---

# Semi-reserved words

In addition to the list of strictly reserved words, there is a class of words and strings that I am going to call "semi-reserved".
- These are named functions or constants (e.g. `pi`) that you can re-assign if you really wanted to... but already come with important meanings from base R.

Arguably the most important semi-reserved character is `c()`, which we use for concatenation; i.e. creating vectors and binding different objects together.

```{r}
my_vector = c(1, 2, 5)
my_vector
```

Try re-assign `pi` for example...

--

**Bottom line:** Don't use (semi-)reserved characters!

---

# Namespace conflicts

A similar issue crops up when we load two packages, which have functions that share the same name. E.g. Look what happens we load the `dplyr` package.
```{r}
library(stats)
library(dplyr)
```

--

The messages that you see about some object being *masked from 'package:X'* are warning you about a namespace conflict.
- E.g. Both `dplyr` and the `stats` package (which gets loaded automatically when you start R) have functions named "filter" and "lag".

---

# Namespace conflicts (cont.)

The potential for namespace conflicts is a result of the OOP approach.<sup>1</sup>
- Also reflects the fundamental open-source nature of R and the use of external packages. People are free to call their functions whatever they want, so some overlap is only to be expected.

.footnote[
<sup>1</sup> Similar problems arise in virtually every other programming language (Python, C, etc.)
]

--

Whenever a namespace conflict arises, the most recently loaded package will gain preference. So the `filter()` function now refers specifically to the `dplyr` variant.

But what if we want the `stats` variant? Well, we have two options:
1. Temporarily use `stats::filter()`
2. Permanently assign `filter = stats::filter`

---

# Solving namespace conflicts

### 1. Use `package::function()`

We can explicitly call a conflicted function from a particular package using the `package::function()` syntax. For example:
```{r}
stats::filter(1:10, rep(1, 2))
```

--

The `::` syntax also means that we can call functions without loading package first. E.g. As long as `dplyr` is installed on our system, then `dplyr::filter(iris, Species=="virginica")` will work.
  
---

# Solving namespace conflicts (cont.)

### 2. Assign `function = package::function`
  
A more permanent solution is to assign a conflicted function name to a particular package. This will hold for the remainder of your current R session, or until you change it back. E.g.
```{r, eval=F}
filter = stats::filter ## Note the lack of parentheses.
filter = dplyr::filter ## Change it back again.
```

---

# General advice

<!-- I would generally advocate for the temporary `package::function()` solution.  -->

Load your most important packages last. (E.g. Load the tidyverse after you've already loaded any other packages.)
  
Other than that, simply pay attention to any warnings when loading a new package and `?` is your friend if you're ever unsure. (E.g. `?filter` will tell you which variant is being used.)
- Problematic namespace conflicts are rare. But it's good to be aware of them. 
  
---

# User-side namespace conflicts

A final thing to say about namespace conflicts is that they don't only arise from loading packages. They can arise when users create their own functions with a conflicting name.
- E.g. If I was naive enough to create a new function called `c()`.

--

</br>

In a similar vein, one of the most common and confusing errors that even experienced R programmers run into is related to the habit of calling objects "df" or "data"... both of which are functions in base R!
- See for yourself by typing `?df` or `?data`.

Again, R will figure out what you mean if you are clear/lucky enough. But, much the same as with `c()`, it's relatively easy to run into problems.
- Case in point: Triggering the infamous "object of type closure is not subsettable" error message. (See from 1:45 [here](https://rstudio.com/resources/rstudioconf-2020/object-of-type-closure-is-not-subsettable/).)

---

# Indexing

Indexing in R starts at 1 (not 0 as in many PLs such as Python, C, C++, etc.)

Example: Create a vector of 10 numbers using `seq` (use `?` to learn what it does):
  ```{r, eval=FALSE}
  x = seq(1,10,1)
  ```
  - `x[i]` will select the ith element (try it yourself)
  
--
  
  - Indexes can be negative, why?
  
--
  
  - You can also select multiple consecutive elements: `x[begin : end]`

---

# The `$` Operator

Let's create a small data.frame: 
```{r, eval = F}
# Create data.frame
dd = data.frame(x = seq(1,100,10), y = seq(1, 200, 20), z = seq(1,300,30))
# View the data.frame
dd
```

`$` allows you to select df columns:

- `dd$x` will select column `x`
--
- Of course, once you select a column, you can select part of it by using indexes...

--

- but you can also use indexes to select columns, how?

--

- `dd[rows list, columns list]` (note that `dd[1]` selects column 1, `dd[1,]` row 1)

---

# Subsetting rows (with data.frame)

- Select rows where `x` is equal to 1: 

  `dd[dd$x ==1,]`

- Select rows where `x` in some values:

  `dd[dd$x %in% c(11,21,31),]`

- Select rows where `x` is different from 91:

  `dd[dd$x != 91,]`

- Select rows where `x` is greater or equal than 40:

  `dd[dd$x >= 40,]`
---

# More operators

https://www.tutorialspoint.com/r/r_operators.htm

<!-- (let's play with them for a few minutes to get familiar with them) -->

---

# Factors 

- Factors are the data objects which are used to categorize the data and store it as levels

- The function `factor` converts vectors (or variables) to factors
  - You can also use it to reorder the levels (useful for regressions)
  - You can also create levels labels
  
---
# Factors (cont.)

```{r, eval = T}
data <- c("East","West","East","North","North","East","West",
          "West","West","East","North", "South")
# Create the factors
data <- factor(data)
data

# Apply the factor function with required order of the level.
factor(data,levels = c("North", "East","South", "West"))

# Apply the factor function with required order of the level and change labels.
factor(data,levels = c("North", "East","South", "West"), labels = c("N", "E", "S", "W"))

```

---
class: inverse, center, middle
name: intro

# More useful things

---
# Build-in datasets 

- R comes with a lot of (build-in datasets](http://www.sthda.com/english/wiki/r-built-in-data-sets). See also [here](https://stackoverflow.com/questions/33797666/how-do-i-get-a-list-of-built-in-data-sets-in-r) to learn how to get a list of these datasets

- Some packages like **dplyr** also have build-in datasets (mpg, starwars, etc.)

---

# Application Programming Interface (API)

There are tons of very useful API available for R
  - [Census data](https://cran.r-project.org/web/packages/censusapi/vignettes/getting-started.html)
  - [Twitter](https://www.earthdatascience.org/courses/earth-analytics/get-data-using-apis/use-twitter-api-r/)
  - etc.
---

# Reading data files

- There are several functions to read any type of data in R, from Excel files to Stata file to csv...

- But the fastest and easy and fastest way to read almost any file (even compressed ones) is [fread](https://www.rdocumentation.org/packages/data.table/versions/1.14.0/topics/fread) (part of the **data.table** library)
  - Usually, the only argument needed is the file name, `fread` will automatically figure out which file you are trying to load!


---

# Rstudio tips

- RStudio is a pretty powerful text editor with a lot of nice features

- For some tips see: https://appsilon.com/rstudio-shortcuts-and-tips/)